---
title: "HIV Study Dashboard"
format:
  dashboard:
    theme: simplex
execute:
  echo: false
  warning: false
  message: false
---

```{=html}
<style>
.bslib-value-box .bslib-full-screen-enter {
  display: none !important;
}
</style>
```

```{r}
#| include: false

library(httr2)
library(dplyr)
library(purrr)
library(plotly)
library(bslib)
library(readr)

# LabKey helpers ----

base_url <- "https://www.labkey.org"
container_path <- "home/Demos/HIV Study Tutorial"

# Streams the TSV export to disk (req_perform path=) then reads it back, so peak
# memory stays bounded by curl's buffer, not the result-set size. FieldKey gives
# machine-friendly column names; maxRows=-1 lifts LabKey's default 100-row cap.
data_dir <- "data"

lk_export_rows <- function(
  schema_name,
  query_name,
  columns = NULL,
  dest = file.path(data_dir, paste0(query_name, ".tsv"))
) {
  request_url <- paste0(
    base_url,
    "/",
    utils::URLencode(container_path),
    "/query-exportRowsTsv.api"
  )

  dir.create(dirname(dest), showWarnings = FALSE, recursive = TRUE)

  request(request_url) |>
    req_url_query(
      schemaName = schema_name,
      `query.queryName` = query_name,
      `query.columns` = columns,
      headerType = "FieldKey",
      `query.maxRows` = -1
    ) |>
    req_perform(path = dest)

  read_tsv(dest, show_col_types = FALSE)
}

# Formatting helpers ----

prettify_label <- function(x) {
  x |>
    gsub("_", " ", x = _) |>
    gsub("([a-z])([A-Z])", "\\1 \\2", x = _) |>
    tools::toTitleCase()
}

first_numeric_col <- function(df) {
  df |>
    select(where(is.numeric)) |>
    select(-matches("participant|visit|date", ignore.case = TRUE)) |>
    names() |>
    pluck(1)
}

# Plotting ----

cohort_colors <- c(
  "No Treatment" = "#C1652E",
  "DRV" = "#2E5C8A",
  "ETR" = "#7A8C42"
)

plot_mean_trend <- function(df, demographics) {
  y_col <- first_numeric_col(df)

  # Append the participant count to each cohort's legend entry, e.g. "DRV (n=3)".
  cohort_n <- demographics |>
    distinct(ParticipantId, cohort) |>
    count(cohort, name = "n")

  label_for <- function(cohort_name) {
    n <- cohort_n$n[cohort_n$cohort == cohort_name]
    if (length(n) == 0) cohort_name else sprintf("%s (n=%d)", cohort_name, n)
  }

  # Re-key the color vector onto the labelled names so colors stay consistent.
  labeled_colors <- setNames(
    unname(cohort_colors),
    vapply(names(cohort_colors), label_for, character(1))
  )

  summary_df <- df |>
    mutate(date = as.Date(date), log_value = log10(.data[[y_col]] + 1)) |>
    inner_join(demographics, by = "ParticipantId") |>
    summarise(
      mean_log = mean(log_value, na.rm = TRUE),
      se_log = sd(log_value, na.rm = TRUE) / sqrt(sum(!is.na(log_value))),
      .by = c(cohort, date)
    ) |>
    mutate(
      ci_lo = mean_log - 1.96 * se_log,
      ci_hi = mean_log + 1.96 * se_log
    ) |>
    arrange(cohort, date)

  # Lookup lets each row pull another cohort's mean on the same date
  by_cohort <- split(summary_df[c("date", "mean_log")], summary_df$cohort)

  lookup_log <- function(cohort_name, on_date) {
    cohort_data <- by_cohort[[cohort_name]]
    if (is.null(cohort_data)) {
      return(NA_real_)
    }
    val <- cohort_data$mean_log[cohort_data$date == on_date]
    if (length(val) == 0) NA_real_ else val[1]
  }

  # Converts a log-mean difference back to a fold-change ratio
  format_fold <- function(log_self, log_other, other_name) {
    if (is.na(log_self) || is.na(log_other)) {
      return(NA_character_)
    }
    ratio <- 10^(log_self - log_other)
    if (ratio >= 1) {
      sprintf("%.1fx higher than %s", ratio, other_name)
    } else {
      sprintf("%.1fx lower than %s", 1 / ratio, other_name)
    }
  }

  comparisons <- list(
    "No Treatment" = c("DRV", "ETR"),
    "DRV" = c("No Treatment", "ETR"),
    "ETR" = c("No Treatment", "DRV")
  )

  summary_df <- summary_df |>
    rowwise() |>
    mutate(hover = {
      others <- comparisons[[cohort]]
      fold_lines <- vapply(
        others,
        \(other) format_fold(mean_log, lookup_log(other, date), other),
        character(1)
      )
      fold_lines <- fold_lines[!is.na(fold_lines)]

      stat_lines <- c(
        sprintf(
          "Log10 mean: %.2f (95%% CI: %.2f–%.2f)",
          mean_log,
          ci_lo,
          ci_hi
        ),
        sprintf(
          "≈ %s copies/mL",
          format(round(10^mean_log - 1), big.mark = ",")
        )
      )
      paste(c(stat_lines, fold_lines), collapse = "<br>")
    }) |>
    ungroup() |>
    mutate(cohort_label = vapply(cohort, label_for, character(1)))

  summary_df |>
    plot_ly(
      x = ~date,
      y = ~mean_log,
      color = ~cohort_label,
      colors = labeled_colors,
      legendgroup = ~cohort_label
    ) |>
    add_ribbons(
      ymin = ~ci_lo,
      ymax = ~ci_hi,
      line = list(width = 0),
      opacity = 0.4,
      showlegend = FALSE,
      hoverinfo = "skip"
    ) |>
    add_lines(
      text = ~hover,
      hovertemplate = paste0(
        "%{x|%b %d, %Y}<br>",
        "%{text}",
        "<extra>%{fullData.name}</extra>"
      )
    ) |>
    layout(
      xaxis = list(title = "Visit Date"),
      yaxis = list(
        title = paste0("Log10(", prettify_label(y_col), "), mean (95% CI)")
      ),
      legend = list(orientation = "h", x = 0, y = -0.25),
      margin = list(l = 70, r = 20, t = 30, b = 70)
    )
}

# Value box helpers ----

darken_hex <- function(hex, factor = 0.65) {
  rgb_vals <- grDevices::col2rgb(hex) * factor
  grDevices::rgb(rgb_vals[1], rgb_vals[2], rgb_vals[3], maxColorValue = 255)
}

stat_box <- function(cohort_name) {
  color <- cohort_colors[[cohort_name]]

  note_tag <- htmltools::tags$div(
    style = "color:#000;",
    "copies/mL at latest visit"
  )

  value_box(
    title = htmltools::tags$span(
      style = sprintf("font-weight:700;color:%s;", darken_hex(color)),
      cohort_name
    ),
    value = htmltools::tags$span(
      style = sprintf("color:%s;", color),
      fmt_geo(cohort_name)
    ),
    note_tag,
    theme = "light",
    height = "175px",
    fill = FALSE,
    full_screen = FALSE
  )
}
```

```{r}
#| include: false

# Load data ----

demographics <- lk_export_rows(
  "study",
  "Demographics",
  columns = "ParticipantId,cohort"
) |>
  filter(cohort != "HIVNeg")

viral_pcr <- lk_export_rows("study", "ViralLoad_PCR")

# Render pulls fresh, so render time is the data timestamp.
data_pulled_at <- format(Sys.time(), "%Y-%m-%d %H:%M UTC", tz = "UTC")

# Summary stats ----

y_col <- first_numeric_col(viral_pcr)

# Join to the HIVNeg-filtered demographics so counts cover only shown cohorts.
counted <- viral_pcr |>
  inner_join(demographics, by = "ParticipantId")

cohort_counts <- counted |>
  summarise(
    n_participants = n_distinct(ParticipantId),
    n_obs = n(),
    .by = cohort
  )

visits_per_participant <- counted |>
  count(ParticipantId, name = "n_visits")

overall_counts <- tibble::tibble(
  n_obs = nrow(counted),
  n_participants = nrow(visits_per_participant),
  median_visits = stats::median(visits_per_participant$n_visits)
)

latest_by_cohort <- viral_pcr |>
  inner_join(demographics, by = "ParticipantId") |>
  mutate(date = as.Date(date)) |>
  filter(date == max(date, na.rm = TRUE), .by = cohort) |>
  summarise(
    geo_mean = 10^(mean(log10(.data[[y_col]] + 1), na.rm = TRUE)) - 1,
    .by = cohort
  )

# Formatting helpers ----

fmt_geo <- function(cohort_name) {
  val <- latest_by_cohort$geo_mean[latest_by_cohort$cohort == cohort_name]
  if (length(val) == 0 || is.na(val)) {
    return("N/A")
  }
  format(round(val), big.mark = ",")
}
```

Synthetic LabKey data comparing two antiretroviral regimens against an untreated arm. Participants with HIV receiving no treatment maintained a substantially higher viral load since Janurary 2021 than either treated cohort.

## Row {height="92%"}

### Column {width="70%"}

```{r}
#| title: "Viral Load by Cohort Over Time (PCR)"
plot_mean_trend(viral_pcr, demographics)
```

### Column {width="30%"}

```{r}
#| expandable: false

stat_box("No Treatment")
```

```{r}
#| expandable: false

stat_box("DRV")
```

```{r}
#| expandable: false

stat_box("ETR")
```

## Row {height="8%" .flow}

Data: LabKey public [HIV Study Tutorial](https://www.labkey.org/home/Demos/HIV%20Study%20Tutorial/project-begin.view), `r overall_counts$n_obs` viral-load measurements from `r overall_counts$n_participants` participants (median `r overall_counts$median_visits` visits each). Last refreshed `r data_pulled_at`.
